---
title: "svm_gibbs_final"
output:
  word_document: default
  html_document: default
  pdf_document: default
date: "2024-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval = FALSE}
# Install relevant packages
library(devtools)
install_github("olafmersmann/truncnorm", force = TRUE)
```

# Import the dataset

```{r}
# Load data 
library(quantmod)

# Import data
data <- getSymbols('^GSPC', src='yahoo', auto.assign=FALSE)

# Get returns
returns <- diff(log(Cl(data)))
returns <- returns[-1]
returns <- as.vector(returns)

# Store snp data
snp_data <- as.data.frame(data)
time <- row.names(snp_data)

# Clean up
rm(data)
```

```{r}
# Store length of returns vector and initial parameter values
T <- length(returns)
initial <- list(h = rep(0,T), sigma2 = 0.02, phi = 0.95, mu = 0)
```

# Preliminaries for Gibbs Sampling Algorithm 

## Defining the hyperparameters

```{r}
# Hyper-parameters for ht sampler 
nu2 <- initial$sigma2 / (1 + initial$phi^2); tau <- initial$sigma2 / (1 - initial$phi^2)

# Hyper-parameters for sigma2 sampler
alpha_sigma <- 2.5; beta_sigma <- 0.025

# Hyper-parameters for phi sampler 
alpha_phi <- 0; beta2_phi <- 1 

# Hyper-parameters for mu sampler 
alpha_mu <- 0; beta2_mu <- 100

# List storing all hyper-parameters
hyperparameters <- list(h = c(nu2, tau), sigma2 = c(alpha_sigma, beta_sigma),
                        phi = c(alpha_phi, beta2_phi), mu = c(alpha_mu, beta2_mu))
```

## Function for sampling ht

```{r}
# Function for sampling ht
sample_ht <- function(data, hyperparameters, h, sigma, phi, mu) {
  
  ## Parameters needed
  T <- length(data)
  
  ## Updated h_1
  h[1] <- rnorm(1, mean = mu, sd = sqrt(hyperparameters[2]))

  ## Update h_[2:T-1]
  for (t in 2:(T-1)) {
    
    ### Parameters for proposal distribution
    hstar <- mu + (phi * (h[t-1] - mu + h[t+1] - mu)) / (1 + phi^2)
    psi <- hstar + (hyperparameters[1] / 2) * (data[t]^2 * exp(-hstar) - 1)

    ### Rejection Sampling 
    repeat{
      proposal <- rnorm(1, psi, hyperparameters[2])
      u <- runif(1, min = 0, max = 1)
      
      #### Acceptance probability
      log_fstar <- (-1 / 2) * proposal - (data[t]^2 / 2) * (exp(-proposal))
      log_gstar <- (-1 / 2) * proposal - (data[t]^2 / 2) * (exp(-hstar) * (1 + hstar) - 
                                                           proposal * exp(-hstar))
      prob <- exp(log_fstar - log_gstar)
      
      #### Decide whether to keep or discard sample
      if (u <= prob) {
        h[t] <- proposal
        break # Exit the repeat loop after a successful sample
      }
    }
  }
  
  ## Return updated h vector
  return(h)
}

# Function for sampling sigma2 
sample_sigma2 <- function(data, hyperparameters, h, phi, mu) {
  
  ## Parameters for inverse gamma distribution
  T <- length(data); ht <- h[-T]; ht_plus1 <- h[-1] 
  shape_sigma2 <- hyperparameters[1] + T/2
  scale_sigma2 <- hyperparameters[2] + (1/2) * (h[1] - mu)^2 * (1 - phi^2) 
                    + (1/2) * sum((ht_plus1 - mu - phi * (ht - mu))^2)
  
  ## Draw sigma2 
  sigma2 <- 1 / rgamma(1, shape = shape_sigma2, rate = scale_sigma2)
  return(sigma2)
}

# Load necessary library 
library(truncnorm)
# Function for sampling phi
sample_phi <- function(data, hyperparameters, h, sigma2, mu) {
  
  ## Parameters for truncated normal distribution 
  T <- length(data); ht <- h[-T]; ht_plus1 <- h[-1] 
  var_phi <- (1 / hyperparameters[2] + (-(h[1] - mu)^2 
                                            + sum((ht - mu)^2)) / sigma2)^-1
  mean_phi <- var_phi * (hyperparameters[1] / hyperparameters[2] 
                         + sum((ht_plus1 - mu) * (ht - mu)) / sigma2)
  
  ## Draw phi  
  phi <- rtruncnorm(1, a=-1, b=1, mean = mean_phi, sd = sqrt(var_phi))
  return(phi)
}

# Function for sampling mu 
sample_mu <- function(data, hyperparameters, h, sigma2, phi) {
  
  ## Parameters for normal distribution
  T <- length(data); ht <- h[-T]; ht_plus1 <- h[-1] 
  var_mu <- (1 / hyperparameters[2] + (1 - phi^2 + (T - 1) * 
                                            (1 - phi)^2) / sigma2)^-1
  mean_mu <- var_mu * (hyperparameters[1] / hyperparameters[2] 
                       + (h[1] * (1 - phi^2) + (1 - phi) * sum(ht_plus1 - phi * ht)) / sigma2)
  
  ## Draw mu 
  mu <- rnorm(1, mean = mean_mu, sd = sqrt(var_mu))
  return(mu)
} 
```

# Gibbs Sampling Algorithm

```{r}
# Gibbs sampling 
gibbs_sampling <- function(data, hyperparameters, initial.values, iterations) {
  
  ## Initialize Parameters
  T <- length(data)
  h <- initial.values$h
  sigma2 <- initial.values$sigma2
  phi <- initial.values$phi
  mu <- initial.values$mu
  hyperparameters_h <- hyperparameters$h
  hyperparameters_sigma2 <- hyperparameters$sigma2
  hyperparameters_phi <- hyperparameters$phi
  hyperparameters_mu <- hyperparameters$mu

  # Initialize the matrix to store sampled values
  # +3 for sigma2, phi, and mu
  sampled_values <- matrix(nrow = iterations, ncol = T + 3)
  colnames(sampled_values) <- c(paste("h", 1:T, sep = ""), "sigma2", "phi", "mu")
  
  ## Gibbs Cycle 
  for (g in 1:iterations) {
    # Step 1: Sample h_t
    h <- sample_ht(data, hyperparameters_h, h, sigma2, phi, mu)
    
    # Step 2: Sample sigma2
    sigma2 <- sample_sigma2(data, hyperparameters_sigma2, h, phi, mu)
    
    # Step 3: Sample phi
    phi <- sample_phi(data, hyperparameters_phi, h, sigma2, mu)
    
    # Step 4: Sample mu
    mu <- sample_mu(data, hyperparameters_mu, h, sigma2, phi)
    
    # Save sampled values in the matrix
    sampled_values[g, 1:T] <- h
    sampled_values[g, (T+1):(T+3)] <- c(sigma2, phi, mu)
  }
  
  return(sampled_values)
}
```

# Implementation of Gibbs Sampler on SP500 Data

Note: the Gibbs Sampler takes ~ 2 to 3 minutes to run 

```{r}
set.seed(123)

# Lists needed as Gibbs input
T <- length(returns)
initial <- list(h = rep(0,T), sigma2 = 0.02, phi = 0.95, mu = 0)
iterations <- 12000

# Gibbs sampling
gibbs_sample_full <- gibbs_sampling(returns, hyperparameters, initial, iterations)
```


# Analysis 

## Data Cleaning & Transformations

Note: Run this section only once as it is removing columns. Otherwise, re-run the gibbs sampler before repeating this section.

First we make the full gibbs sample easy to work with.
```{r}
# Convert Gibbs sample matrix into dataframe for easier access
gibbs_sample_full <- as.data.frame(gibbs_sample_full)

# Remove the last h column (not updated)
gibbs_sample_full <- gibbs_sample_full[-(ncol(gibbs_sample_full)-3)]

# Label rows
rownames(gibbs_sample_full) <- paste("Cycle ", 1:(nrow(gibbs_sample_full)))

# Add beta to the sample dataset
gibbs_sample_full$beta <- with(gibbs_sample_full, exp(gibbs_sample_full$mu / 2))

# Export as csv 
#write.csv(gibbs_sample_full, "gibbs-full-sample.csv")
```

Now we discard the burn-in period.
```{r}
# Remove Burn-In Period 
gibbs_sample <- gibbs_sample_full[2001:12000, ]

# View first 6 values of model parameters (use ncol instead of hard-coding b/c data updates every day)
head(gibbs_sample[,c((ncol(gibbs_sample) - 3):(ncol(gibbs_sample)))])
```

## Trace Plots 

First without burn in:

```{r}
par(mfrow = c(1, 3))

# Trace plots of parameters
plot(1:(nrow(gibbs_sample_full)), gibbs_sample_full$sigma2, col = 'firebrick', type = "b", 
     main = "", xlab = "Cycle", ylab = "σ^2 | y", pch = '.')

plot(1:(nrow(gibbs_sample_full)), gibbs_sample_full$phi, col = 'firebrick', type = "b", 
     main = "", xlab = "Cycle", ylab = "ɸ | y", pch = '.')

plot(1:(nrow(gibbs_sample_full)), gibbs_sample_full$mu, col = 'firebrick', type = "b", 
     main = "", xlab = "Cycle", ylab = "μ | y", pch = '.')

#plot(1:(nrow(gibbs_sample)), gibbs_sample_full$beta, col = 'firebrick', type = "b", 
#     main = "", xlab = "Cycle", ylab = "β | y", pch = '.')

mtext("", side = 3, outer = TRUE, padj = 3)

# Reset to default margins
par(mfrow = c(1, 1))  # the default mar values
```

Next with burn in:

```{r}
# Set up the plotting area with adjusted margins
# mar sets the margins as (bottom, left, top, right) in lines
# oma sets the outer margins in lines (bottom, left, top, right)
par(mfrow = c(1, 3))

# Trace plots of parameters
plot(1:(nrow(gibbs_sample)), gibbs_sample$sigma2, col = 'firebrick', type = "b", 
     main = "", xlab = "Cycle", ylab = "σ^2 | y", pch = '.')

plot(1:(nrow(gibbs_sample)), gibbs_sample$phi, col = 'firebrick', type = "b", 
     main = "", xlab = "Cycle", ylab = "ɸ | y", pch = '.')

plot(1:(nrow(gibbs_sample)), gibbs_sample$mu, col = 'firebrick', type = "b", 
     main = "", xlab = "Cycle", ylab = "μ | y", pch = '.')

#plot(1:(nrow(gibbs_sample)), gibbs_sample$beta, col = 'firebrick', type = "b", 
#     main = "", xlab = "Cycle", ylab = "β | y", pch = '.')

mtext("", side = 3, outer = TRUE, padj = 3)

# Reset to default margins
par(mfrow = c(1, 1))  # the default mar values
```

## Summary Statistics and Densities 

```{r}
# Means 
means <- list(sigma2 = mean(gibbs_sample$sigma2), phi = mean(gibbs_sample$phi), mu = mean(gibbs_sample$mu), beta = mean(gibbs_sample$beta))

c(means$sigma2, means$phi, means$mu, means$beta)

# Variance
var <- list(sigma2 = var(gibbs_sample$sigma2), phi = var(gibbs_sample$phi), mu = var(gibbs_sample$mu), beta = var(gibbs_sample$beta))

c(var$sigma2, var$phi, var$mu, var$beta)
```

```{r}
# Set up the plotting area
par(mfrow = c(1, 3))

# Densities of parameters
hist(gibbs_sample$sigma2, freq = FALSE, xlim = c(0, 4e-05), ylim = c(0, 2e+05), 
     col = 'indianred', main = "", xlab = "σ^2 | y")

hist(gibbs_sample$phi, freq = FALSE, xlim = c(0.96, 1), ylim = c(0, 300), 
     col = 'indianred', main = "", xlab = "ɸ | y")

hist(gibbs_sample$mu, freq = FALSE, xlim = c(-9.5,-8.5), ylim = c(0, 10),
     col = 'indianred', main = "", xlab = " μ| y")

#hist(gibbs_sample$beta, freq = FALSE, xlim = c(0.01, 0.012), ylim = c(0, 1000),
#     col = 'indianred', main = "", xlab = " β| y")

mtext("", side = 3, outer = TRUE, padj = 3)

# Reset to default
par(mfrow = c(1, 1))
```

## ACF Plots

```{r}
# Set up the plotting area
par(mfrow = c(1, 3))

acf(ts(gibbs_sample$sigma2), lag.max = 80, ylab = "ACF for σ^2", main = "", col = 'firebrick')
acf(ts(gibbs_sample$phi), lag.max = 80, ylab = "ACF for ɸ", main = "", col = 'firebrick')
acf(ts(gibbs_sample$mu), lag.max = 80, ylab = "ACF for μ", main = "", col = 'firebrick')
#acf(ts(gibbs_sample$beta), lag.max = 40, ylab = "ACF for β", main = "", col = 'firebrick')

mtext("", side = 3, outer = TRUE, padj = 3)

# Reset to default
par(mfrow = c(1, 1))
```

## Implied Volatility 
Note: Run this section only once as it is removing columns. Otherwise, re-run the gibbs sampler before repeating this section.

```{r}
# Calculate the means of each of the first 4355 variables
rm_col <- 20 # set number of ht you want to remove from end b/c of high variance in mean
mean.h <- colMeans(gibbs_sample[, 1:(ncol(gibbs_sample) - (rm_col + 4))], na.rm = TRUE) 
exp.mean.h <- exp(mean.h/2)

# Assuming the row names in 'data' are dates you want to use as labels
x_vals <- time[1:length(exp.mean.h)]

# Creating a numeric sequence for plotting
index <- 1:length(x_vals)

# Plotting using numeric indices but labeling with time values
par(mar=c(5, 4, 4, 2) + 0.1, mgp=c(2, 0.5, 0))
plot(index, exp.mean.h, type = "l", main = "Implied Volatility Over Time",
     xlab = "Time", ylab = "Volatility, exp(h_t/2)", col = "firebrick", lwd = 2, axes = FALSE, xaxt = 'n')

# Add y-axis
axis(2)

# Enhance with grid
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted")

# Adding a box around the plot
box()

# Selecting exactly four indices for x-axis labels
label_indices <- round(seq(from = 1, to = length(x_vals), length.out = 4))
label_dates <- x_vals[label_indices]

# Adding custom x-axis with four time labels, horizontally aligned
axis(1, at = label_indices, labels = label_dates, las = 1, cex.axis = 0.7)
```